Automatic extraction of symbolic network signatures from malware code

### 2016-12-15 paper.tex
Reorganizing and final paper

\begin{abstract}
We propose the automatic extraction of symbolic network signatures from malware code to aid and automate network-based malware detection systems.
\end{abstract}

\section{Motivation}
Malware, short for malicious software, 
is one of the primary threats to modern-day networks, especially
with the mass proliferation of the Internet. These are used to 
disrupt computer operations, extract sensitive information, and 
gain access to private systems.  Infected hosts can participate 
in distributed denial-of-service attacks (DDoS), these class of 
attacks are very hard to detect and mitigate, leading to immense
loss of revenues. A new shift is towards creating malware for 
profit called botnets, in which a customer can buy a network of
infected hosts for a price. In such a setting, the malware is
specifically designed to evade detection, thus motivating the need
for better detection tools for malware.

The most prevalent form of malware detection is endhost-based detection
(commonly known as anti-virus software). Detected malware's code is hashed
and the code signatures are uploaded to a central repository. By maintaining
an up-to-date repository of malware signatures, the endhost software can
detect malware by comparing program signatures with the repository. However, 
endhost-based malware detection can fail due to various reasons~\cite{networksig}:
\begin{itemize}
	\item \textbf{Anti-virus software is not operational or up to date.} 
	Due to this, the endhost machine can be infected by malware without any
	noticeable effects, thus the user will be oblivious of the infection.
	\item \textbf{Packer programs.} To render programs stealthy, malware authors
	employ packer programs~\cite{packer}. Packers change the program content so that
	its signature differs but its functionality has not changed, thus avoiding
	detection by endhost signature-based software.
\end{itemize}

@@

\section{Challenges}
We abstract the problem to the following: given a program point,
can we automatically provide an ordered set of symbolic packets 
that must be received (and sent) by the client such that the program 
reaches the program point. The symbolic packets can be used as
network signatures and used to detect malware (a set of packets
satisfying the symbolic packets in order will infer that there 
is a malware with high confidence). By constructing symbolic packet
signatures, we can generalise the intrusion policy and reduce 
true negatives (thus not affecting legitimate traffic).

@@

We model a malware program as a infinite reactive system 
of the form:
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
}
\end{lstlisting}
Each iteration of the loop, the program receives a
packet, and based on current state and packet received,
it performs certain actions. Using static analysis
techniques, we can identify two set of variables, \emph{state}
and \emph{packet} variables. State variables represent the 
program state in terms of the protocol, which packet variables
are used to represent the current packet being processed by
the program.
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
	if (a == 1 and packet[8] = "b") {
		a = a + 1;
	}
	...
}
\end{lstlisting}	
In the above code snippet, $a$ is a state variable
and $packet$ is a packet variable. This also
illustrates the packet processing 
function of the program, which based on the current
input packet and state variables, either transitions
to a new program state or performs an action (attack or
send packets). 
\begin{lstlisting}	
a = 0;
while(True) {
packet = socket.receive();
	if (a == 0 and packet[0] == "a") {
		a = a + 1;
	}
	if (a == 1 and packet[0] == "b") {
		a = a + 1;
	}
	if (a == 2 and packet[0] == "c") {
		attack()  [the attack point is provided as 
				input to analysis]
	}	
}
\end{lstlisting}	
The above code snippet represents the complete model
of the malware program, where based on input packets and
current state, the program transitions to a new state. 
%
%The symbolic automaton for this program is 
%shown in \Cref{fig:progautomaton}. Network signatures
%for the malware are all paths from start state ($a=0$)
%to state of attack. In this case, the sequence of packets
%in order to initiate an attack is $a* \rightarrow b* \rightarrow c*$.
%\begin{figure}[h]
%	\centering
%	\begin{tikzpicture}[shorten >=0.5pt,node distance=3cm,on grid,auto] 
%	\node[state, initial] (v0)   {$a = 0$}; 
%	\node[state] (v1) [right=of v0] {$a = 1$}; 
%	\node[state] (v2) [right=of v1] {$a = 2$};
%	\node[state] (v3) [right=of v2] {$attack$}; 	
%	\path[->] 
%	(v0) edge node {$a*$} (v1)
%	(v1) edge node {$b*$} (v2)
%	(v2) edge node {$c*$} (v3);
%	\end{tikzpicture}
%	\caption{Symbolic automaton for malware program.}
%	\label{fig:progautomaton}
%\end{figure}

\section{Related Work}
An important assumption that we make is that of the simplification of the botnet client code to a set of \textbf{if (state, packet)} statements. Looking at existing research, significant work has gone into automatic protocol reverse-engineering of botnet C\&C protocols using dynamic binary analysis. This generally follows a two step process: Extract the protocol grammar that captures the structure of messages comprising the protocol. Knowing the protocol message format would enable the construction of the protocol state machine, which represents the sequence of messages as a specification of protocol states and transition between states (based on messages received).

@@

We can model the malware state transition 
by the following constraints: 
\begin{equation}
\forall i. 0 \leq i \leq n-2.~\forall (\gamma, \delta) \in I. ~\gamma(S_i, P_{i}) \Rightarrow S_{i+1} \Leftrightarrow \delta(S_i, P_{i})
\end{equation}

@@

Also, this procedure only returns a particular signature
of length $n$. We can generalize it by analyzing the 
transitions of $M$ to generalize the signature. For
example, if the $i^{th}$ packet only matches certain 
predicates in the transition, we can generalize the packet on
these bits (the rest of the bits are don't care). To 
generate the different singatures of length $n$, 
we can add blocking clauses for the signature we obtained 
to find a signature different from the current one.

\subsection{Example}
Let us consider the following program receiving two-bit packets:
\begin{lstlisting}	
s1 = 0;
s2 = 0;
while(True) {
packet = socket.receive();
if (s1 == 0 and packet[0] == 0) {
s1 = 1;
}
if (s1 == 1 and s2 == 0 and packet[1] == 1) {
s2 = (not s2) and packet[1];
}
if (s1 == 1 and s2 == 1 and packet[0] == 1 and packet[1] == 1) {
attack()  
}	
}
\end{lstlisting}	
We denote the state variables by $s_1, s_2$ and packet 
variables by $p_1, p_2$.
For the first \texttt{if} block, 
$\gamma: \neg s_1 \wedge \neg p_1$, and $\delta: s_1' \wedge s_2' \Leftrightarrow s_2$ ($s_1$ 
is set by the action). Similarily,
for the second block: 
$\gamma: s_1 \wedge \neg s_2 \wedge p_2$, and $\delta: 
s_1'\Leftrightarrow s_1 \wedge s_2' \Leftrightarrow \neg s_2 \wedge p_1$.
For this program, initial state $S_0 = (0, 0)$.
The attack point predicate $\gamma_M$ is as follows:
\begin{equation}
s_1 \wedge s_2 \wedge p_1 \wedge p_1
\end{equation}
Therefore, if we add the constraints as described in the 
earlier section, for $n=3$, we would obtain a
valid network signature.

@@

Another simplifying assumption is our malware program model 
as a set of if blocks in a while loop. We found real malware programs with
large codebases with mutliple files, thus packet processing can be 
across several procedures, requiring inter-procedural analyses.
Also, it may be difficult to find malware code in the wild, only binaries,
so extraction of signatures for a binary would be useful.

>>>>

\begin{abstract}
Malware is one of the primary threats to modern-day networks. While the most prevalent form of malware detection is endhost-based (e.g., matching code hashes), a detection mechanism integrated directly into the network offers a much-needed additional layer of protection. Towards this goal, we propose automatic generation of accurate network signatures of malware  using program analysis techniques and SMT solvers. Informally, a network signature is the set of messages received by the malware to initiate attacks. These network signatures can be used by intrusion detection systems to detect malware traffic in the network. We present some preliminary work towards this goal. 
\end{abstract}

\section{Motivation}
Malware is software designed to inflict illegitimate action and affect data, hosts or networks. These are used to 
disrupt computer operations, extract sensitive information, and 
gain access to private systems. Infected hosts can participate 
in distributed denial-of-service attacks (DDoS), and these class of 
attacks are very hard to detect and mitigate, leading to immense
loss of revenues. A new shift is towards creating malware for 
profit called botnets, in which a customer can buy a network of
infected hosts for a price. In such a setting, the malware is
specifically designed to evade detection, thus motivating the need
for better detection tools for malware.

The most prevalent form of malware detection is endhost-based detection
(commonly known as anti-virus software). The detected malware's code is hashed and the code signatures are uploaded to a central repository. By maintaining an up-to-date repository of malware signatures, the endhost software can detect malware by comparing program signatures with the repository. However, endhost-based malware detection can fail due to various reasons~\cite{networksig}:
\begin{itemize}
	\item \textbf{Anti-virus software is not operational or up to date.} 
	Due to this, the endhost machine can be infected by malware without any
	noticeable effects, thus the user will be oblivious of the infection.
	\item \textbf{Packer programs.} To render programs stealthy, malware authors
	employ packer programs~\cite{packer}. Packers change the program content so that
	its signature differs but its functionality has not changed, thus avoiding
	detection by endhost signature-based software.
\end{itemize}

@@

\section{Challenges}
We abstract the problem to the following: given a program point,
can we automatically provide an ordered set of symbolic packets 
that must be received (and sent) by the client such that the program 
reaches the program point. The symbolic packets can be used as
network signatures and used to detect malware (a set of packets
satisfying the symbolic packets in order will infer that there 
is a malware with high confidence). By constructing symbolic packet
signatures, we can generalize the intrusion policy and reduce 
true negatives (thus not affecting legitimate traffic).

@@

We model a malware program as a infinite reactive system 
of the form:
\begin{lstlisting}	
while(True) {
	packet = socket.receive();
	...
}
\end{lstlisting}
Each iteration of the loop, the program receives a
packet, and based on current state and packet received,
it performs certain actions. Using static analysis
techniques, we can identify two set of variables, \emph{state}
and \emph{packet} variables. State variables represent the 
program state in 	terms of the protocol, which packet variables
are used to represent the current packet being processed by
the program.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{code.PNG}
	\label{fig:code1}
\end{figure}
In the above code snippet, $a$ is a state variable
and $packet$ is a packet variable. This also
illustrates the packet processing 
function of the program, which based on the current
input packet and state variables, either transitions
to a new program state or performs an action (attack or
send packets). 
\begin{figure}[h]
	\includegraphics[width=\columnwidth]{code2.PNG}
	\label{fig:code2}
\end{figure}
The above code snippet represents the complete model
of the malware program, where based on input packets and
current state, the program transitions to a new state. 
%
%The symbolic automaton for this program is 
%shown in \Cref{fig:progautomaton}. Network signatures
%for the malware are all paths from start state ($a=0$)
%to state of attack. In this case, the sequence of packets
%in order to initiate an attack is $a* \rightarrow b* \rightarrow c*$.
%\begin{figure}[h]
%	\centering
%	\begin{tikzpicture}[shorten >=0.5pt,node distance=3cm,on grid,auto] 
%	\node[state, initial] (v0)   {$a = 0$}; 
%	\node[state] (v1) [right=of v0] {$a = 1$}; 
%	\node[state] (v2) [right=of v1] {$a = 2$};
%	\node[state] (v3) [right=of v2] {$attack$}; 	
%	\path[->] 
%	(v0) edge node {$a*$} (v1)
%	(v1) edge node {$b*$} (v2)
%	(v2) edge node {$c*$} (v3);
%	\end{tikzpicture}
%	\caption{Symbolic automaton for malware program.}
%	\label{fig:progautomaton}
%\end{figure}

\section{Related Work}
An important assumption that we make is that of the simplification of the botnet client code to a set of \textbf{if (state, packet)} statements. Looking at existing research, significant work has gone into automatic protocol reverse-engineering of botnet C\&C protocols using dynamic binary analysis. A general process followed is as follows: Extract the protocol grammar that captures the structure of messages comprising the protocol. Knowing the protocol message format would enable the construction of the protocol state machine, which represents the sequence of messages as a specification of protocol states and transition between states (based on messages received).

@@

We can model the malware state transition 
by the following constraints:

$\forall i. 0 \leq i \leq n-2.~\forall (\gamma, \delta) \in I.~$
\begin{equation}
\gamma(S_i, P_{i}) \Rightarrow S_{i+1} \Leftrightarrow \delta(S_i, P_{i})
\end{equation}

@@

Also, this procedure only returns a particular signature
of length $n$. We can generalize it by analyzing the 
transitions of $M$ to generalize the signature. For
example, if the $i^{th}$ packet only matches certain 
predicates in the transition, we can generalize the packet on
these bits (the rest of the bits are don't care). To 
generate the different signatures of length $n$, 
we can add blocking clauses for the signature we obtained 
to find a signature different from the current one.
\subsection{Example}
Let us consider a program receiving two-bit packets.

We denote the state variables by $s_1, s_2$ and packet 
variables by $p_1, p_2$.
For the first \texttt{if} block, 
$\gamma: \neg s_1 \wedge \neg p_1$, and $\delta: s_1' \wedge s_2' \Leftrightarrow s_2$ ($s_1$ 
is set by the action). Similarly,
for the second block: 
$\gamma: s_1 \wedge \neg s_2 \wedge p_2$, and $\delta: 
s_1'\Leftrightarrow s_1 \wedge s_2' \Leftrightarrow \neg s_2 \wedge p_1$.
\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{code7.PNG}
	\label{fig:code7}
\end{figure}
For this program, initial state $S_0 = (0, 0)$.
The attack point predicate $\gamma_M$ is as follows:
\begin{equation}
s_1 \wedge s_2 \wedge p_1 \wedge p_1
\end{equation}
Therefore, if we add the constraints as described in the 
earlier section, for $n=3$, we would obtain a
valid network signature.

@@

Another simplifying assumption is our malware program model 
as a set of if blocks in a while loop. We found real malware programs with
large codebases with multiple files, thus packet processing can be 
across several procedures, requiring inter-procedural analyses.
Also, it may be difficult to find malware code in the wild, only binaries,
so extraction of signatures for a binary would be useful.
